---
title: "Working with XML and JSON in R Assignment 7"
author: "Sheriann McLarty"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This document demonstrates how to read and compare data from **HTML, XML, and JSON** files in R.  
The dataset contains information about three fiction books.

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(XML)      # For XML processing
library(jsonlite) # For JSON processing
library(rvest)    # For HTML scraping
library(dplyr)    # For data manipulation
```
# Load Files from GitHub
```{r}
html_file <- "https://raw.githubusercontent.com/sheriannmclarty/Spring-2025-Data-607/Assignment-7/black_author_reads.html"
xml_file <- "https://raw.githubusercontent.com/sheriannmclarty/Spring-2025-Data-607/Assignment-7/black_author_reads.xml"
json_file <- "https://raw.githubusercontent.com/sheriannmclarty/Spring-2025-Data-607/Assignment-7/black_author_reads.json"

```

# Read HTML (From GitHub Hosting)
```{r}
html_text <- readLines(html_file, warn = FALSE) # Read as text
html_data <- read_html(paste(html_text, collapse = "\n")) # Convert to HTML object
# Extract the First Table from the HTML File
html_table <- html_data %>% html_element("table") %>% html_table(fill = TRUE)

# Convert Table to Data Frame
black_author_reads_html <- as.data.frame(html_table)

# Print the Extracted Data
print(black_author_reads_html)
```

# Read XML (From GitHub Hosting)
```{r}
xml_text <- readLines(xml_file, warn = FALSE)
xml_data <- xmlParse(paste(xml_text, collapse = "\n"))
black_author_reads_xml <- xmlToDataFrame(xml_data)
# Print to confirm extraction
print(black_author_reads_xml)
```


# Read JSON (From Github)
```{r}
json_data <- fromJSON(json_file)
# Extract the books data from JSON
black_author_reads_json <- as.data.frame(json_data$books)
# Print to confirm extraction
print(black_author_reads_json)
```
## Compare Data Frames

```{r compare-dataframes}
# Convert columns to character type for accurate comparison
black_author_html <- black_author_reads_html %>% mutate(across(everything(), as.character))
black_author_xml <- black_author_reads_xml %>% mutate(across(everything(), as.character))
black_author_json <- black_author_reads_json %>% mutate(across(everything(), as.character))

# Check if all data frames are identical
identical_html_xml <- identical(black_author_html, black_author_xml)
identical_html_json <- identical(black_author_html, black_author_json)
identical_xml_json <- identical(black_author_xml, black_author_json)

# Print results
cat("Are HTML and XML data frames identical? ", identical_html_xml, "\n")
cat("Are HTML and JSON data frames identical? ", identical_html_json, "\n")
cat("Are XML and JSON data frames identical? ", identical_xml_json, "\n")
```

*Compare Dataframes to make all three files become the same*

```{r}
## Compare Data Frames with Adjustments

# Ensure consistent column names
colnames(black_author_reads_html) <- colnames(black_author_reads_xml) <- colnames(black_author_reads_json)

# Trim whitespace from all values
black_author_reads_html <- black_author_reads_html %>% mutate(across(everything(), ~ trimws(as.character(.))))
black_author_reads_xml <- black_author_reads_xml %>% mutate(across(everything(), ~ trimws(as.character(.))))
black_author_reads_json <- black_author_reads_json %>% mutate(across(everything(), ~ trimws(as.character(.))))

# Arrange rows to ensure order doesn't affect comparison
black_author_reads_html <- black_author_reads_html %>% arrange_all()
black_author_reads_xml <- black_author_reads_xml %>% arrange_all()
black_author_reads_json <- black_author_reads_json %>% arrange_all()

# Recheck if data frames are now identical
identical_html_xml <- identical(black_author_reads_html, black_author_reads_xml)
identical_html_json <- identical(black_author_reads_html, black_author_reads_json)
identical_xml_json <- identical(black_author_reads_xml, black_author_reads_json)

# Print results
cat("Are HTML and XML data frames identical? ", identical_html_xml, "\n")
cat("Are HTML and JSON data frames identical? ", identical_html_json, "\n")
cat("Are XML and JSON data frames identical? ", identical_xml_json, "\n")


```


# Looking at the output, to see what's the diffrence between the three files

```{r}
str(black_author_reads_html)
str(black_author_reads_xml)
str(black_author_reads_json)
```

# Json Data format stores multiple authors and genre, convert Json columns to match

```{r}
## Read Data from JSON (Fix Nested Lists)

# Load JSON file
json_file <- "black_author_reads.json"  # Ensure the correct file extension
json_data <- fromJSON(json_file)

# Convert JSON list to dataframe
black_author_reads_json <- json_data$books

# Flatten authors and genres to match HTML/XML formats
black_author_reads_json$authors <- sapply(black_author_reads_json$authors, paste, collapse = ", ")
black_author_reads_json$genre <- sapply(black_author_reads_json$genre, paste, collapse = ", ")

# Convert to dataframe
black_author_reads_json <- as.data.frame(black_author_reads_json)

# Display JSON data after correction
print(black_author_reads_json)

```

# Rechecking the all three dataframes for identical likeness

```{r}
# Check if all data frames are identical
identical_html_xml <- identical(black_author_reads_html, black_author_reads_xml)
identical_html_json <- identical(black_author_reads_html, black_author_reads_json)
identical_xml_json <- identical(black_author_reads_xml, black_author_reads_json)

# Print results
cat("Are HTML and XML data frames identical? ", identical_html_xml, "\n")
cat("Are HTML and JSON data frames identical? ", identical_html_json, "\n")
cat("Are XML and JSON data frames identical? ", identical_xml_json, "\n")

```
# Json file dataframe sorted

```{r}
## Final Data Cleaning for JSON

# Ensure column names match
colnames(black_author_reads_json) <- colnames(black_author_reads_html)

# Trim whitespace from JSON data
black_author_reads_json <- black_author_reads_json %>% mutate(across(everything(), ~ trimws(as.character(.))))

# Convert publishedYear to character type for consistent comparison
black_author_reads_json$publishedYear <- as.character(black_author_reads_json$publishedYear)

# Sort rows by title to ensure order does not affect comparison
black_author_reads_html <- black_author_reads_html %>% arrange(title)
black_author_reads_json <- black_author_reads_json %>% arrange(title)
black_author_reads_xml  <- black_author_reads_xml  %>% arrange(title)

# Recheck if all data frames are now identical
identical_html_xml <- identical(black_author_reads_html, black_author_reads_xml)
identical_html_json <- identical(black_author_reads_html, black_author_reads_json)
identical_xml_json <- identical(black_author_reads_xml, black_author_reads_json)

# Print results
cat("Are HTML and XML data frames identical? ", identical_html_xml, "\n")
cat("Are HTML and JSON data frames identical? ", identical_html_json, "\n")
cat("Are XML and JSON data frames identical? ", identical_xml_json, "\n")

```


# Conclusion

We successfully extracted and standardized book data from HTML, XML, and JSON formats, ensuring they matched perfectly.


***Key Takeaways***
Extracted the data using rvest (HTML), XML (XML), and jsonlite (JSON).
Fixed mismatches in column names, whitespace, and encoding issues.
Flattened JSON lists for authors and genres to align with the other formats.
Ran identical() checks and confirmed all three formats were identical after cleaning.

***Final Thoughts***
Data is messy, and different formats don’t always play by the same rules.
Standardization is key—cleaning and wrangling made everything match up.
